---
title: "ErmineJ GO analysis for H. winteri"
output: html_notebook
---

Load packages.

```{r, setup, include=F}
library(tidyverse)
library(ermineR)
setwd("~/Documents/Alice_winterii/")
```

Load the data, subset the annotation to the 10,447 transcripts included in the gene expression data.
```{r, data}
# outliers
lfmm_out <- read_table("lfmm.txt")
lfmm_out <- lfmm_out %>% 
  mutate(chr = str_replace_all(chr, 'HA', 'Ha'),
         chr = str_replace_all(chr, 'CHR', 'Chr'),
         snp_id = paste(chr, pos, sep = ":"))

pcadapt_out <- read_table("pcadapt.txt")
pcadapt_out <- pcadapt_out %>% 
  filter(PC != 2) %>% # only keep the SNPs associated with PCs 1,3,4
  mutate(chr = str_replace_all(chr, 'HA', 'Ha'),
         chr = str_replace_all(chr, 'CHR', 'Chr'),
         snp_id = paste(chr, pos, sep = ":"))

# genes
genes_out <- read_table("Alice_snps_10kb_genes.tsv")

lfmm_genes <- genes_out %>% 
  filter(snp_id %in% lfmm_out$snp_id) %>% 
  distinct()

pcadapt_genes <- genes_out %>% 
  filter(snp_id %in% pcadapt_out$snp_id) %>% 
  distinct()

# annotation
annotation <- read_tsv("GO_database_result.txt", col_select = 1:2, col_names = c("gene_id", "GO_term"))

annotation <- annotation %>% 
  group_by(gene_id) %>% 
  mutate(GOs = paste(GO_term, collapse =",")) %>% 
  select(-GO_term) %>% 
  distinct()
```

Prep the annotation for the analysis. The github for the R wrapper says "If you want to use your own GO annotations instead of getting files provided by Pavlidis Lab, you can use `makeAnnotation` after turning your annotations into a list. [See the example.](https://github.com/PavlidisLab/ermineR?tab=readme-ov-file#using-your-own-go-annotations)."

So, make a column into a list.

```{r, GO}
go <- as.list(strsplit(annotation$GOs,split=","))
names(go) <- as.character(annotation$gene_id)

go_annotation <- makeAnnotation(go)

```

There are several ways to do a GO enrichment analysis. ORA (over-representation analysis) takes a list of genes above some threshold and asks what they have in common relative to the other genes. The others use ranks, and I don't think we can easily do that with our data. We'll take the list of genes for each subset (lfmm, pcadapt, and for their combo).

```{r ora_lfmm}
lfmm_ora <- ora(hitlist = lfmm_genes$gene_id,
              annotation = go_annotation,
              aspects = c("Molecular Function", "Cellular Component", "Biological Process"),
              pAdjust = "FDR")

lfmm_ora$results %>% arrange(CorrectedPvalue)

```

Two terms significant after FDR with q < 0.1. Now pcadapt!

```{r ora_pcadapt}
pcadapt_ora <- ora(hitlist = pcadapt_genes$gene_id,
              annotation = go_annotation,
              aspects = c("Molecular Function", "Cellular Component", "Biological Process"),
              pAdjust = "FDR")

pcadapt_ora$results %>% arrange(CorrectedPvalue)

```

Nothing is significant after FDR correction but the lowest p-value term is the same as the LFMM set: "modified amino acid metabolic process". The only other one close to it is "glutathione metabolic process", which is marginally significant after p-value correction.

Are the pcadapt top terms just driven by the shared genes?
```{r ora_pcadapt2}
pcadapt_genes_nolfmm <- pcadapt_genes %>% 
  filter(!(gene_id %in% lfmm_genes$gene_id))

pcadapt_ora2 <- ora(hitlist = pcadapt_genes_nolfmm$gene_id,
              annotation = go_annotation,
              aspects = c("Molecular Function", "Cellular Component", "Biological Process"),
              pAdjust = "FDR")

pcadapt_ora2$results %>% arrange(CorrectedPvalue)

```
Yes, the two top terms don't show up here and nothing is significant.

We can look at the combined terms just for fun:

```{r ora_both}
both_genes <- unique(c(lfmm_genes$gene_id, pcadapt_genes$gene_id))

both_ora <- ora(hitlist = both_genes,
              annotation = go_annotation,
              aspects = c("Molecular Function", "Cellular Component", "Biological Process"),
              pAdjust = "FDR")

both_ora$results %>% arrange(CorrectedPvalue)

```

Very similar to pcadapt alone (not surprising since that is most of the genes and they overlap), non-signficant.

Looking at the overlap of 11 in detail, this might just be due to the fact that genes near the SNPs tend to be paralogs (three paired genes and one triplicate paralog near three different shared SNPs). Let's pull out the annotation for just the set of "outlier" genes.

```{r outlierGO}
pcadapt_ann <- annotation %>% 
  filter(gene_id %in% pcadapt_genes_nolfmm$gene_id) %>% 
  arrange(gene_id) %>% 
  mutate(outlier = "selection")

lfmm_ann <- annotation %>% 
  filter(gene_id %in% lfmm_genes$gene_id) %>% 
  arrange(gene_id) %>% 
  mutate(outlier = "climate")

out_ann <- rbind(pcadapt_ann,lfmm_ann) %>% 
  arrange(gene_id)

out_ann

write_csv(out_ann, file="outlier_annotations.csv")
```

